tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-21 paper.tex
Update perf

% \subsection{Phenotype}
% The final phenotype simulation is done in \texttt{tstrait}
% by imposing some environmental noise.
% We assume environmental noise is normally distributed with mean zero
% and a variance of
% \[
% V_G(1-h^2)/{h^2}
% \]
% where $V_G$ is the additive genetic variance from the simulated effect sizes,
% and $h^2$ is the narrow-sense heritability parameter provided as input.
% Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
% and narrow-sense heritability as parameters:
% \begin{lstlisting}[language=Python]
% phenotype_df = tstrait.sim_env(
%     genetic_result.genetic, h2=0.3)
% \end{lstlisting}
% This simulation returns a dataframe
% describing the simulated genetic value, environmental noise and
% resulting phenotype for each individual at each causal site.

% As discussed above, there are significant advantages to providing a modular
% API. For convenience, we also provide a single
% \texttt{sim\_phenotype} function that combines the process of simulating
% effect sizes, generating genetic values and imposing environmental noise.

\subsection{Implementation and validation}
\texttt{Tstrait} is written entirely in Python. Many numerical operations are
peformed using standard array-oriented approaches and therefore
highly efficient~\citep{numpy}. Generating the genetic values for a given
trait is computationally intensive. This is done by a
straightforward tree traversal algorithm, which is
accelerated using the \texttt{numba} JIT compiler for Python~\citep{numba} yielding
excellent performance.

The \texttt{tstrait} codebase includes an extensive suite of unit tests,
which are automatically run as part of the development process. The
output of \texttt{tstrait} has been statistically validated against
[FILL ME IN].

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
For comparison, we also report the CPU time required by a
modified version of the code used by \citet{martin2017} for a
subset of these simulated ARGs.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the VCFs involved are very large, and these tools assume that
the entire dataset fits in RAM.
Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
modest for all these simulations.

% TODO this is to be rewritten with just a single example from FC
% dataset and optionally also mentioning real data from 1000G (space
% permitting).
Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
realistic data. For example, \texttt{tstrait} is used to simulate traits from
2.7 million individuals in the French Canadian dataset, which is the largest
known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
practical to simulate quantitative traits from this largest ever simulated
dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
also used to measure the computational performance.

>>>>

\subsection{Implementation and validation}
\texttt{Tstrait} is written entirely in Python. Numerical operations
are either peformed using standard array-oriented
approaches~\citep{numpy} or accelerated
using the \texttt{numba} JIT compiler~\citep{numba}.
The \texttt{tstrait} codebase includes an extensive suite of unit tests,
which are automatically run as part of the development process. The
output of \texttt{tstrait} has been extensively validated against
theoretical expectations, as well as the output
of \texttt{AlphaSimR}~\citep{gaynor2021} and
\texttt{simplePHENOTYPES}~\citep{fernandes2020}.

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
modest for all these simulations.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the VCFs involved are very large, and these tools assume that
the entire dataset fits in RAM.
To emphasise this point, we also generated
quantitative traits for the simulations based on
the French-Canadian pedigree~\citet{anderson2023}, discussed in
the introduction. It took 60 seconds and XXX GB of RAM to
simulate a trait with 1000 causal sites using \texttt{tstrait}.
Using one byte per genotype, the full variant matrix of this
dataset would require 140TB of RAM.
% If we have space, mention the real-data sims

% Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
% realistic data. For example, \texttt{tstrait} is used to simulate traits from
% 2.7 million individuals in the French Canadian dataset, which is the largest
% known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
% simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
% practical to simulate quantitative traits from this largest ever simulated
% dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
% also used to measure the computational performance.
